---
title: "Quality Control"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In this document we are going to perform quality control on the cells.

## Setup

Set chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/01-quality-control.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required packages:

```{r pacman}
pacman::p_load(
  AnnotationHub,
  BiocSingular,
  devtools,
  eulerr,
  patchwork,
  readr,
  robustbase,
  scales,
  scater,
  scran
)
```

Read experiment data:

```{r}
sce <- readRDS("data/quant.rds")
```

## Exploration

We begin by exploring the dataset which contains `r ncol(sce)` cells and `r nrow(sce)` genes.

## Metrics

Identify mitochondrial transcripts:

```{r}
hub <- AnnotationHub()

ens <- hub[["AH78811"]]

ids <- mapIds(ens, keys = rowData(sce)$gene_id, keytype = "GENEID", column = "SEQNAME")

mit <- which(ids == "MT")
```

Compute QC metrics for each cell:

```{r}
sce <- addPerCellQC(sce, subsets = list(MT = mit))
```

Inspect the distributions of QC metrics:

```{r, fig.height = 9, fig.width = 10}
wrap_plots(
  plotColData(sce, "sum") + scale_y_log10() + ggtitle("Total count"),
  plotColData(sce, "detected") + scale_y_log10() + ggtitle("Detected features"),
  plotColData(sce, "subsets_MT_percent") + ggtitle("MT percent"),
  plotColData(sce, "altexps_ERCC_percent") + ggtitle("ERCC percent")
)
```

### Dimensionality reduction {.tabset}

Calculate dimenionality reduction data:

```{r}
set.seed(110111011)
sizeFactors(sce) <- librarySizeFactors(sce)
sce <- logNormCounts(sce)
sce <- runPCA(sce, BSPARAM = ExactParam())
sce <- runTSNE(sce, dimred = "PCA")
sce <- runUMAP(sce, dimred = "PCA")
```

Inspect dimensionality reduction plots:

#### PCA

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotPCA(sce, colour_by = "sum") + ggtitle("Total count"),
  plotPCA(sce, colour_by = "detected") + ggtitle("Detected features"),
  plotPCA(sce, colour_by = "subsets_MT_percent") + ggtitle("MT percent"),
  plotPCA(sce, colour_by = "altexps_ERCC_percent") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### TSNE

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotTSNE(sce, colour_by = "sum") + ggtitle("Total count"),
  plotTSNE(sce, colour_by = "detected") + ggtitle("Detected features"),
  plotTSNE(sce, colour_by = "subsets_MT_percent") + ggtitle("MT percent"),
  plotTSNE(sce, colour_by = "altexps_ERCC_percent") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### UMAP

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotUMAP(sce, colour_by = "sum") + ggtitle("Total count"),
  plotUMAP(sce, colour_by = "detected") + ggtitle("Detected features"),
  plotUMAP(sce, colour_by = "subsets_MT_percent") + ggtitle("MT percent"),
  plotUMAP(sce, colour_by = "altexps_ERCC_percent") + ggtitle("ERCC percent"),
  ncol = 2
)
```

## Cell Filtering

We will now perform filtering to remove low-quality cells:

### Manual threshold {.tabset}

The first approach to identifying low-quality cells is to apply manual thresholds on the QC metrics:

```{r}
dat <- DataFrame(
  low_lib_size = sce$sum < 1e5,
  low_n_features = sce$detected < 5e3,
  high_subsets_MT_percent = sce$subsets_MT_percent > 10,
  high_altexps_ERCC_percent = sce$altexps_ERCC_percent > 10
)

sce$manual_discard <- dat$low_lib_size | dat$low_n_features | dat$high_subsets_MT_percent | dat$high_altexps_ERCC_percent
```

Inspect distribution of QC metrics:

```{r}
wrap_plots(
  plotColData(sce, "sum", colour_by = "manual_discard") + scale_y_log10() + ggtitle("Total count"),
  plotColData(sce, "detected", colour_by = "manual_discard") + scale_y_log10() + ggtitle("Detected features"),
  plotColData(sce, "subsets_MT_percent", colour_by = "manual_discard") + ggtitle("MT percent"),
  plotColData(sce, "altexps_ERCC_percent", colour_by = "manual_discard") + ggtitle("ERCC percent")
)
```

Inspect dimensionality reduction plots:

#### PCA

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotPCA(sce, colour_by = "sum", shape_by = "manual_discard") + ggtitle("Total count"),
  plotPCA(sce, colour_by = "detected", shape_by = "manual_discard") + ggtitle("Detected features"),
  plotPCA(sce, colour_by = "subsets_MT_percent", shape_by = "manual_discard") + ggtitle("MT percent"),
  plotPCA(sce, colour_by = "altexps_ERCC_percent", shape_by = "manual_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### TSNE

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotTSNE(sce, colour_by = "sum", shape_by = "manual_discard") + ggtitle("Total count"),
  plotTSNE(sce, colour_by = "detected", shape_by = "manual_discard") + ggtitle("Detected features"),
  plotTSNE(sce, colour_by = "subsets_MT_percent", shape_by = "manual_discard") + ggtitle("MT percent"),
  plotTSNE(sce, colour_by = "altexps_ERCC_percent", shape_by = "manual_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### UMAP

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotUMAP(sce, colour_by = "sum", shape_by = "manual_discard") + ggtitle("Total count"),
  plotUMAP(sce, colour_by = "detected", shape_by = "manual_discard") + ggtitle("Detected features"),
  plotUMAP(sce, colour_by = "subsets_MT_percent", shape_by = "manual_discard") + ggtitle("MT percent"),
  plotUMAP(sce, colour_by = "altexps_ERCC_percent", shape_by = "manual_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

### Adaptive threshold {.tabset}

The next approach is to identify cells that are outliers based on the median absolute deviation for the QC metrics:

```{r}
dat <- quickPerCellQC(sce, percent_subsets=c("subsets_MT_percent", "altexps_ERCC_percent"))

sce$adaptive_discard <- dat$discard
```

Inspect distribution of QC metrics:

```{r}
wrap_plots(
  plotColData(sce, "sum", colour_by = "adaptive_discard") + scale_y_log10() + ggtitle("Total count"),
  plotColData(sce, "detected", colour_by = "adaptive_discard") + scale_y_log10() + ggtitle("Detected features"),
  plotColData(sce, "subsets_MT_percent", colour_by = "adaptive_discard") + ggtitle("MT percent"),
  plotColData(sce, "altexps_ERCC_percent", colour_by = "adaptive_discard") + ggtitle("ERCC percent")
)
```

Inspect dimensionality reduction plots:

#### PCA

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotPCA(sce, colour_by = "sum", shape_by = "adaptive_discard") + ggtitle("Total count"),
  plotPCA(sce, colour_by = "detected", shape_by = "adaptive_discard") + ggtitle("Detected features"),
  plotPCA(sce, colour_by = "subsets_MT_percent", shape_by = "adaptive_discard") + ggtitle("MT percent"),
  plotPCA(sce, colour_by = "altexps_ERCC_percent", shape_by = "adaptive_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### TSNE

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotTSNE(sce, colour_by = "sum", shape_by = "adaptive_discard") + ggtitle("Total count"),
  plotTSNE(sce, colour_by = "detected", shape_by = "adaptive_discard") + ggtitle("Detected features"),
  plotTSNE(sce, colour_by = "subsets_MT_percent", shape_by = "adaptive_discard") + ggtitle("MT percent"),
  plotTSNE(sce, colour_by = "altexps_ERCC_percent", shape_by = "adaptive_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### UMAP

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotUMAP(sce, colour_by = "sum", shape_by = "adaptive_discard") + ggtitle("Total count"),
  plotUMAP(sce, colour_by = "detected", shape_by = "adaptive_discard") + ggtitle("Detected features"),
  plotUMAP(sce, colour_by = "subsets_MT_percent", shape_by = "adaptive_discard") + ggtitle("MT percent"),
  plotUMAP(sce, colour_by = "altexps_ERCC_percent", shape_by = "adaptive_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

### Outlier threshold {.tabset}

The last approach is to identify outliers in high-dimensional space based on the QC metrics for each cell:

```{r}
mat <- cbind(log10(sce$sum), log10(sce$detected), sce$subsets_MT_percent, sce$altexps_ERCC_percent)
adj <- adjOutlyingness(mat, only.outlyingness = TRUE)
out <- isOutlier(adj, type = "higher")
sce$outlier_discard <- out
```

Inspect distribution of QC metrics:

```{r}
wrap_plots(
  plotColData(sce, "sum", colour_by = "outlier_discard") + scale_y_log10() + ggtitle("Total count"),
  plotColData(sce, "detected", colour_by = "outlier_discard") + scale_y_log10() + ggtitle("Detected features"),
  plotColData(sce, "subsets_MT_percent", colour_by = "outlier_discard") + ggtitle("MT percent"),
  plotColData(sce, "altexps_ERCC_percent", colour_by = "outlier_discard") + ggtitle("ERCC percent")
)
```

Inspect dimensionality reduction plots:

#### PCA

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotPCA(sce, colour_by = "sum", shape_by = "outlier_discard") + ggtitle("Total count"),
  plotPCA(sce, colour_by = "detected", shape_by = "outlier_discard") + ggtitle("Detected features"),
  plotPCA(sce, colour_by = "subsets_MT_percent", shape_by = "outlier_discard") + ggtitle("MT percent"),
  plotPCA(sce, colour_by = "altexps_ERCC_percent", shape_by = "outlier_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### TSNE

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotTSNE(sce, colour_by = "sum", shape_by = "outlier_discard") + ggtitle("Total count"),
  plotTSNE(sce, colour_by = "detected", shape_by = "outlier_discard") + ggtitle("Detected features"),
  plotTSNE(sce, colour_by = "subsets_MT_percent", shape_by = "outlier_discard") + ggtitle("MT percent"),
  plotTSNE(sce, colour_by = "altexps_ERCC_percent", shape_by = "outlier_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

#### UMAP

```{r, fig.height = 9, fig.width = 10}
patchwork::wrap_plots(
  plotUMAP(sce, colour_by = "sum", shape_by = "outlier_discard") + ggtitle("Total count"),
  plotUMAP(sce, colour_by = "detected", shape_by = "outlier_discard") + ggtitle("Detected features"),
  plotUMAP(sce, colour_by = "subsets_MT_percent", shape_by = "outlier_discard") + ggtitle("MT percent"),
  plotUMAP(sce, colour_by = "altexps_ERCC_percent", shape_by = "outlier_discard") + ggtitle("ERCC percent"),
  ncol = 2
)
```

### Comparison

Compare the number of cells selected between methods using an Euler diagram:

```{r}
mat <- cbind(Keep = TRUE, Manual = sce$manual_discard, Adaptive = sce$adaptive_discard, Outlier = sce$outlier_discard)
fit <- euler(mat, shape = "circle")
col <- RColorBrewer::brewer.pal(n = 4, name = "Pastel2")
plot(fit, fill = col, quantities = TRUE)
```

### Selection

We are going to use the adaptive filtering to remove low quality cells:

```{r}
sce <- sce[, !sce$adaptive_discard]
```

## Gene filtering

Compute QC metrics for each gene:

```{r}
sce <- addPerFeatureQC(sce)

dat <- as.data.frame(rowData(sce))
```

### Low abundance

Plot the distribution of log-means across all genes:

```{r}
ggplot(dat, aes(mean)) + 
  geom_histogram(bins = 100, colour = "black", fill = "grey") + 
  scale_x_log10(labels = comma) +
  theme_bw()
```

Define low abundance genes using a mean threshold:

```{r}
dat$low_abundance <- dat$mean < 1
```

Tabulate the number of low abundance genes:

```{r}
knitr::kable(table(dat$low_abundance), col.names = c("Low abundance", "Count"))
```

### Low frequency

Plot the percentange of expressing cells against the log-mean expression for each gene:

```{r}
ggplot(dat, aes(mean, detected)) +
  geom_point(alpha = 0.1, colour = "grey") +
  geom_smooth(colour = "black", se = FALSE) +
  scale_x_log10(labels = comma) +
  theme_bw()
```

Define low frequency genes as having a minimum of 1 read in at least 1 percent of cells:

```{r}
dat$low_frequency <- !rowSums(counts(sce) >= 1) >= (ncol(sce) * 0.01)
```

Tabulate the number of low frequency genes:

```{r}
knitr::kable(table(dat$low_frequency), col.names = c("Low frequency", "Count"))
```

### Comparison

Compare the number of genes selected between methods using an Euler diagram:

```{r}
mat <- cbind(Keep = TRUE, Mean = dat$low_abundance, Freq = dat$low_frequency)
fit <- euler(mat, shape = "circle")
col <- RColorBrewer::brewer.pal(n = 3, name = "Pastel2")
plot(fit, fill = col, quantities = TRUE)
```

### Filter

We are going to use the frequency-based filter as it tends to be less aggressive:

```{r}
sce <- sce[!dat$low_frequency, ]
```

## Summary

After quality control we have a dataset with `r ncol(sce)` cells and `r nrow(sce)` genes.

### Output

```{r}
saveRDS(sce, "data/01-quality-control.rds")
```

### Session

Print version information:

```{r session}
session_info()
```
