---
title: "Cell cycle assignment"
author: "James Ashmore"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Define chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/07-cell-cycle.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Attach required packages:

```{r pacman}
pacman::p_load(
  SingleR,
  scran,
  scRNAseq,
  scater,
  org.Mm.eg.db,
  here,
  readr
)
```

Import experiment data:

```{r}
rds <- here("data/06-doublet-detection.rds")

sce <- read_rds(rds)

alt <- altExp(sce, "original", withColData = TRUE)
```

## Expression

Inspect the relative expression of cyclins across the clusters:

```{r}
idx <- grep("^Ccn[abde][0-9]$", rowData(alt)$gene_name)

ids <- rownames(alt)[idx]

plotHeatmap(alt, features = sort(ids), order_columns_by = "cluster", cluster_rows = FALSE)
```

Inspect the relative expession of cell-cycle related genes across the clusters:

```{r}
sel <- select(org.Mm.eg.db, keys = "GO:0007049", keytype = "GOALL", column = "ENSEMBL")

idx <- rowData(alt)$gene_id %in% sel$ENSEMBL

ids <- rownames(alt)[idx]

plotHeatmap(alt, features = ids, order_columns_by = "cluster", show_rownames = FALSE)
```

## Reference {.tabset}

Obtain and normalize the Buettner ESC data:

```{r}
ref <- BuettnerESCData()

ref <- logNormCounts(ref)
```

Use reference data to assign phase to test data:

```{r}
ids <- Reduce(intersect, list(rownames(ref), rowData(alt)$gene_id, sel$ENSEMBL))

res <- pairwiseWilcox(logcounts(ref), ref$phase, direction = "up", subset.row = ids)

top <- getTopMarkers(res$statistics, res$pairs)

rownames(alt) <- rowData(alt)$gene_id

fit <- SingleR(test = alt, ref = ref, label = ref$phase, genes = top)

sce$phase <- factor(fit$labels, levels = c("G1", "G2M", "S"))
```

Inspect the distribution of cell-cycle phases on reduced dimensions:

### PCA

```{r}
plotPCA(sce, colour_by = "phase", text_by = "cluster")
```

### TSNE

```{r}
plotTSNE(sce, colour_by = "phase", text_by = "cluster")
```

### UMAP

```{r}
plotUMAP(sce, colour_by = "phase", text_by = "cluster")
```

## Summary

### Output

Write experiment data:

```{r}
write_rds(sce, here("data/07-cell-cycle.rds"))
```

### Session

Print session information:

```{r session}
devtools::session_info()
```
